### 2013-08-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding final edits to spn chapter

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin, control rod, and moderator materials and
the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
while a cross section of the materials configuration along the
horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
control rods were removed for these problems leaving only moderator in
their place. A detailed view of the assembly bottom is given in
Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
vacuum conditions are used as well as on the top and right boundaries
in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized control rods. Each fuel pin is resolved by a $2
      \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is homogenized
      moderator/control rods and darker purple/red material is
      homogenized moderator/fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is homogenized moderator/control rods and darker
      purple/red material is homogenized moderator/fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

>>>>

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and the $SP_N$ discretization
in a criticality calculation. A cross section perpendicular to the
vertical axis showing homogenized fuel pin, control rod, and moderator
materials and the associated grid is given by
Figure~\ref{fig:problem3_radial_mat} while a cross section of the
materials configuration perpendicular to the horizontal axis is given
by Figure~\ref{fig:problem3_axial_mat}. The control rods were removed
for these problems leaving only moderator in their place. A detailed
view of the assembly bottom is given in
Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
vacuum conditions are used as well as on the top and right boundaries
in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of the entire assembly. For the spatial discretization,
each fuel pin (gray regions in Figure~\ref{fig:problem3_radial_mat})
is resolved by a $2 \times 2$ mesh with materials and cross sections
homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized control rods. Each fuel pin is resolved by a $2
      \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is homogenized
      moderator/control rods and darker purple/red material is
      homogenized moderator/fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is homogenized moderator/control rods and darker
      purple/red material is homogenized moderator/fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 total
histories for this problem) with the number of iterations required to
converge increasing as a function of spectral radius. Near a spectral
radius of 0.97, the number of histories required to converge MCSA in
less than 100 iterations takes a dramatic rise that exhibits neither
exponential nor power law behavior. As the spectral radius approaches
1, the number of histories required becomes significant and
effectively impractical to compute. Even with this simple diffusion
problem, the behavior is consistent with that observed for the fuel
assembly problem with $SP_1$ discretization. In that case, we
estimated a spectral radius of $\approx 0.998$, larger than any of the
spectral radii that could be computed within even 90 minutes of
compute time for this simple two dimensional problem. For that large
of a spectral radius, even 1,000,000 histories ($\approx 50$ per DOF)
was not enough to provide convergence. Even if single solve times of
an hour can be tolerated, dozens of solves are typically required to
compute the multiplication factor and flux spectrum within the $SP_N$
eigenvalue scheme for more difficult problems like the fuel assembly,
making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}
